---
title: "Reading "
lightbox: true
format:
  html:
    toc: true
  pdf:
    documentclass: article
---

# Exercise 1: grouping and summarizing

- Load the `data_MPA`
- Create a new object called `MPA_summary` that has the mean biomass by period and treatment status
- When and where is the biomass highest?

Hint: use a combination of `group_by()` and `summarize()`

```{r}
# My code will be here
```

**Advanced** (optional): Use the mean values stored in `MPA_summary` to calculate the difference-in-differences like we did last week

```{r}
# My code will be here
```


# Exercise 2: Obtaining data from the wild

How to find the data:
1. Go to [iattc.org](https://www.iattc.org/)
2. In the top menu, **hover** over `DATA`
3. Click on "Public domain"
4. You will be taken to a page titled "Public domain data for download"
5. We will use 
6. Click on `CatchByFlagGear.zip` to prompt a download
7. Save the zip file to inside your EVR628 project at: `data/raw/`
8. Using your finder / explorer, navigate to `EVR628/data/raw/` and unzip the `CatchByFlagGear.zip` file
9. You will get a new folder called 

# Exercise 3: Reading data

```{r}
# My code will be here
```

# Exercise 4: Manipulating data

```{r}
# My code will be here
```
